---
title: "Data Challenge 2 (Data Preparation)"
author: "YOUR NAME"
date: "DATE"
output: html_document
---

# Introduction (remove this part when submitting)

We're going to produce a data report in RMarkdown. Think of RMarkdown as Microsoft Word, but then much cooler (automizable!).

R and RMarkdown allow you to combine inline code snippets (the part between ```{r}```) with text in markdown (e.g., `#### Exercise 1`). For now, it's important to realize that once you click on the "Knit" button, this `.Rmd` file is rendered into a HTML page. 

As you'll see, this creates headers, separates markdown cells from code blocks, makes links clickable, and many other things. Have a watch at this video [tutorial](https://www.youtube.com/watch?v=tKUufzpoHDE) or play round with this [cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) to learn the basics. 

Please keep in mind that your notebook only runs if there are no errors. In case you suspect there might be an error somewhere, you may want to add `eval=FALSE` to `{r}` so that it skips that block and still renders your file.

## Goals

The goal of this data challenge is to create a data pipeline that imports the data, performs a series of transformation steps, and outputs the data so that you can share it with others (e.g., potential collaborators on a team project).

It contains three stages:

1) Code necessary to read in the data and perform some basic cleaning steps (input)
2) Apply one or more preprocessing steps (e.g., merge, group, rank, counter, string splitting) (transformation), and
3) An exploration of interesting relationships in the data. Spot interesting patterns, visualize the relationships and exports your results in an appropriate format (e.g., `.csv` or `.jpg` for plots).

## Ideas for data exploration

Here are a couple of ideas you may want to look into, but feel more than welcome to add your own ones!

* What are the most common genre tag words? Do they differ between countries? And continents? Which artist has appeared most frequently in the top 200? 

* Did our music preferences change since the start of the corona pandemic? (tip: make a time-series plot!)

* Can you find a general trend in how track rankings fluctuate over time? For how many weeks do most songs remain in the top 200 on average? How often did songs lose their top 200 ranking and climb up to the top 10 again? Is there a relationship between the ranking and the popularity score? 

* What are non-obvious predictors of the number of streams for a song? (tip: use a linear model: `lm(streams ~ ... + ... + ..., data=df)`)

* Of course, you can come up with your own ideas here!


\

## 1. Input

```{r eval=FALSE}

# ADD ALL OF THE REQUIRED CODE TO LOAD THE DATA AND PERFORM SOME INITIAL CLEANING STEPS THAT ARE REQUIRED TO RUN THIS DOCUMENT
streams <- read.csv("streams.csv", sep=",")
songs <- read.csv("songs.csv", sep=",")
country_codes <- read.csv("country_codes.csv", sep=";")

# add more...
```

\ 

## 2. Transformation

We perform the following steps in our data preparation pipeline:

```{r}

# e.g., merge

```

Another transformation step...

```{r}
# e.g., filter
```

...And another one

```{r}
# e.g., group
```

## 3. Output

### Idea 1

```{r}
# E.g., export of a plot
```

### Idea 2

```{r}
# E.g., export of a regression table
```

### ...
