---
title: "Data Preparation"
author: "Data Challenge 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*In this tutorial we're going to explore some of the data sets underlying the paper that estimates the COVID-19 impact on music consumption ([Sim et al. 2020](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3649085)). Along the way, you'll practice the skills you have been equipped with in the DataCamp tutorials.*


--- 

### Prerequisites
* [DataCamp Introduction to Tidyverse](https://campus.datacamp.com/courses/introduction-to-the-tidyverse/data-wrangling-1?ex=1) (chapter 1 and 3)
* [DataCamp Cleaning Data in R](https://www.datacamp.com/courses/free-introduction-to-r) (chapter 1, 2)


--- 

### Learning Objectives
* Apply data wrangling in tidyverse and dplyr (filter, sort, group, summarize)
* Conduct cleaning operations (convert data type, trim string, remove duplicates)
* Write Markdown documentation

--- 


### Submission 
Students are expected to hand in a digital copy of their answers and code according to the [submission guidelines](https://dprep.hannesdatta.com/#submission-guidelines). 


--- 

### Support Needed? 
For technical issues outside of scheduled classes, please check the [support section](https://dprep.hannesdatta.com/docs/course/support) on the course website.


---

### 1. Merging Data
...

```{r load_data}
 #mobility <- read.csv("2020_NL_Region_Mobility_Report.csv", sep=",")
```
\
**Exercise 1**     
...

**Solution**  
...

\

**Exercise 2**  
... 

**Solution**  
...
\

---

### 2. Feature Engineering

\
**Exercise 3**     
...

**Solution**  
...

\

**Exercise 4**  
... 

**Solution**  
...
\


### 3. Model Building

\
**Exercise 3**     
...

**Solution**  
...

\

**Exercise 4**  
... 

**Solution**  
...
\




### 4. Writing Markdown Documentation

...