---
title: "Data Challenge 2"
author: "YOUR NAME"
date: "DATE"
output: html_document
---

This time, we're going to run our analysis in a R Markdown file which is somewhat similar to a Jupyter Notebook. Both tools allow you to combine inline code snippets with text in markdown. Either watch this video [tutorial](https://www.youtube.com/watch?v=tKUufzpoHDE) or study this [cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) to learn the basics. It is important to realize that your notebook only runs if there are no errors. Hence, if there is an error in one of the R blocks, you can add `eval=FALSE` to `{r}` so that it skips that block and still renders your file.


#### Exercise 1

```{r}
# install.packages('dplyr')  you only need to install the package once!
# library(dplyr)
# your answer
```

#### Exercise 2

```{r}
# your answer
```

#### Exercise 3

```{r}
# your answer
```

#### Exercise 4

```{r}
# your answer
```

#### Exercise 5

```{r}
# your answer
```

#### Exercise 6

```{r}
# your answer
```

#### Exploratory Analysis

```{r}
# your answer
```

