---
title: "Data Challenge 1 (Data Exploration)"
author: "YOUR NAME"
date: "DATE"
output: html_document
---

# Introduction (remove this part when submitting)

We're going to produce a data report in RMarkdown. Think of RMarkdown as Microsoft Word, but then much cooler (automizable!).

R and RMarkdown allow you to combine inline code snippets (the part between ```{r}```) with text in markdown (e.g., `#### Exercise 1`). For now, it's important to realize that once you click on the "Knit" button, this `.Rmd` file is rendered into a HTML page. 

As you'll see, this creates headers, separates markdown cells from code blocks, makes links clickable, and many other things. Have a watch at this video [tutorial](https://www.youtube.com/watch?v=tKUufzpoHDE) or play round with this [cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) to learn the basics. 

Please keep in mind that your notebook only runs if there are no errors. In case you suspect there might be an error somewhere, you may want to add `eval=FALSE` to `{r}` so that it skips that block and still renders your file.

## Goals

The goal of this data challenge is to create a data report that you can share with others (e.g., potential collaborators on a team project).

It contains three stages:

1) Code necessary to read in the data and perform some basic cleaning steps
2) Generating an overview about the data (e.g., summary statistics, report on missingness, number of observations, etc.), and
3) An exploration of interesting relationships in the data. Spot interesting patterns, visualize the relationships and provide a brief write-up of your exploration.

## Ideas for data exploration

Here are a couple of ideas you may want to look into, but feel more than welcome to add your own ones!

- Comparison of mobility patterns in the first and second COVID-19 wave. Did our behavior change throughout time? And how about the summer break? 

- How does your own place of residence relate to the rest of the Netherlands? (e.g., create a ranking!)

- The southern provinces of the Netherlands initially suffered the most from COVID-19 in terms of the number of positive cases (possibly due to [carnival](https://dutchreview.com/news/first-man-to-get-coronavirus-covid-19-in-the-netherlands-celebrated-carnival/)). As such, stricter policies were imposed by the government than in other parts of the country. Can you find patterns in the data that illustrate the (in)effectiveness of these measures? 

- Extract the Google Mobility statistics for *a different country*, and re-process your RMarkdown. Compare the Covid stats for these two countries.

- Incorporate the ideas on Data Delivery and Data Checking, expressed in Mela's (2011) paper on [Data Selection and Procurement](https://pubsonline.informs.org/doi/abs/10.1287/mksc.1110.0650).

- Of course, you can come up with your own ideas here!



\

```{r, echo = FALSE}

# ADD ALL OF THE REQUIRED CODE TO LOAD THE DATA AND PERFORM SOME INITIAL CLEANING STEPS THAT ARE REQUIRED TO RUN THIS DOCUMENT
mobility <- read.csv("2020_NL_Region_Mobility_Report.csv", sep=",")

# add more...
```

\ 

# Overview of the data

This is how the data set looks like.

```{r}
head(mobility)
```

Some in-line text here: The result of `1+1` is (`r 1+1`).

```{r}
# some code
```

...Some more description

```{r}
# some code
```

# Exploration of interesting relationships

## Idea 1

```{r}
# some code
```


## Idea 2

## ...
