---
title: "Tutorial (week 3) - ggplot, GitHub branches & pull requests, and Makefiles"
subtitle: "Data Preparation and Programming Skills (2026/2027)"
author: "Javier Torralba Flores"
date: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
  message: false
format:
  beamer:
    theme: metropolis
    colortheme: seahorse
    code-overflow: wrap
    code-font-size: 0.85em
    aspectratio: 169
    fontsize: 11pt
    highlight-style: tango
    include-in-header:
      text: |
        \usepackage{xcolor}
        \newcommand{\workbg}{\setbeamercolor{background canvas}{bg=yellow!12}}
        \newcommand{\normalbg}{\setbeamercolor{background canvas}{bg=white}}
---

## Tutorial roadmap (90 minutes)

Today follows one project story with four blocks:

1. Opening and orientation - 10 min
2. Coding with `ggplot2` - 30 min
3. Collaboration with Git and GitHub - 30 min
4. Automation with Makefile + wrap-up - 20 min

Total: 90 minutes.

---

## Opening story (10 minutes)

Imagine this timeline:

- Monday: you make a plot.
- Tuesday: your teammate changes style and labels.
- Wednesday: someone asks what changed.
- Next week: data changes, but output does not update.

Today we solve this with one clear workflow.

---

## Learning goals

By the end, you can:

- build and improve plots with `ggplot2`
- use `git status`, `git add`, `git commit`, and `git push`
- work with branches and pull requests
- write one Makefile rule
- explain why outputs can stay old

---

## Section 1: Coding with `ggplot2` (30 minutes)

```{=latex}
\normalbg
```

Question for this block:

Which creators look strong in both reach and quality?

---

## Section 1 setup

Run this code first:

```{r}
library(tidyverse)

if (!file.exists("video_view.csv")) {
  data_url <- paste0(
    "https://raw.githubusercontent.com/hannesdatta/",
    "course-dprep/refs/heads/main/material/tutorials/r-bootcamp-rev/",
    "video_view.csv"
  )
  download.file(data_url, "video_view.csv")
}

videos <- read_csv("video_view.csv")
```

---

## How `ggplot2` works

`ggplot2` uses a simple recipe:

1. pick data
2. map variables to axes and colors (`aes`)
3. pick chart type (`geom_*`)
4. improve labels and style

Template:

```r
ggplot(data = ..., aes(x = ..., y = ...)) +
  geom_...()
```

---

## One dataset, many chart types

Same grammar, different `geom_*`:

- `geom_point()` for relationships
- `geom_histogram()` for distributions
- `geom_col()` for category comparisons
- `geom_line()` for trends over time

This is why ggplot feels consistent.

---

## Mapping vs setting

- Inside `aes(...)`: map data to visual properties
  - Example: `aes(color = creator_id)`
- Outside `aes(...)`: set fixed style
  - Example: `color = "steelblue"`

Mapped color gives a legend. Fixed color does not.

---

## WORK TIME - Exercise 1

```{=latex}
\workbg
```

Build a baseline scatter plot:

1. data: `videos`
2. x-axis: `impressions_n`
3. y-axis: `watch_rate`
4. chart type: `geom_point()`

---

## WORK TIME - Exercise 2 (improve plot + AI)

```{=latex}
\workbg
```

Improve your plot for a non-technical reader:

1. add title and axis labels
2. add color mapping
3. use a readable theme
4. ask AI for one improvement idea

Use this prompt:

```text
I am a beginner in ggplot2.
Please improve this plot for readability.
Add better labels, colors, and legend.
Explain each change in one sentence.
```

Deliverables:

- improved plot
- one AI suggestion you used
- one AI suggestion you did not use

---

## Section 2: Collaboration with Git + GitHub (30 minutes)

```{=latex}
\normalbg
```

Goal: share changes clearly and safely.

---

## Save, commit, pull, pull request

- save: update file on your machine
- commit: save snapshot in Git history
- `git pull`: receive updates from GitHub
- pull request: propose your updates for merge

---

## Branching in simple words

`main` is your stable version.

- create a branch to work safely
- change files on that branch
- commit and push branch
- open pull request
- merge into `main` when ready

Think: work in branch first, merge later.

---

## Section 2 setup

Use any repo you can push to.

If needed, use this practice repo:

`https://github.com/jtorralbaflores/video-view-analysis-dprep`

Edit only `README.md` in this block.

---

## WORK TIME - Exercise 3

```{=latex}
\workbg
```

Run one full branch cycle:

```bash
git pull origin main
git switch -c feature-week3-readme
# edit one line in README.md
git status
git add README.md
git commit -m "Update README for week 3"
git push -u origin feature-week3-readme
```

---

## WORK TIME - Exercise 4

```{=latex}
\workbg
```

Create and merge a pull request on GitHub:

1. open PR from `feature-week3-readme` to `main`
2. check changed files and diff
3. merge pull request
4. delete branch on GitHub

---

## Section 3: Automation with Makefile (20 minutes)

```{=latex}
\normalbg
```

Goal: stop guessing what to rerun.

---

## How one Make rule works

Each rule has three parts:

- target: what to build
- prerequisites: what it depends on
- command: how to build it

If a prerequisite changes, `make` rebuilds target.

---

## Why outputs can stay old

If a prerequisite is missing, `make` cannot see the change.

Then your script changed, but output may not rebuild.

---

## Section 3 setup

Create a fresh folder and enter it.

```bash
mkdir week3_make_practice
```

In that folder, create `build_plot.R` and `makefile`.

---

## WORK TIME - Exercise 5

```{=latex}
\workbg
```

Create two files: `build_plot.R` and `makefile`.

Step 1: write `build_plot.R`.

---

## Exercise 5 - build_plot.R

```{=latex}
\workbg
```

```r
library(tidyverse)

url <- paste0(
  "https://raw.githubusercontent.com/hannesdatta/course-dprep/",
  "refs/heads/main/material/tutorials/r-bootcamp-rev/video_view.csv"
)
if (!file.exists("video_view.csv")) download.file(url, "video_view.csv")

videos <- read_csv("video_view.csv")

ggplot(videos, aes(impressions_n, watch_rate)) +
  geom_point(color = "steelblue", alpha = 0.7)

ggsave("plot_week3.png", width = 7, height = 4)
```

---

## Exercise 5 - makefile and run

```{=latex}
\workbg
```

Step 2: write `makefile` and run `make`.

```make
plot_week3.png: build_plot.R video_view.csv
	Rscript build_plot.R
```

Run:

```bash
make -n
make
make
```

---

## WORK TIME - Exercise 6

```{=latex}
\workbg
```

Part A: change one line in `build_plot.R` (for example point color), then run:

```bash
make -n
make
```

Part B: now break the rule:

```make
plot_week3.png: video_view.csv
	Rscript build_plot.R
```

Edit `build_plot.R` again and run `make`.

What happened? Then fix the rule.

---

## Wrap-up

```{=latex}
\normalbg
```

You now practiced one complete workflow:

- make and improve a plot
- branch, commit, push, and open PR
- automate output with Make
- understand why outputs can stay old

---

## Exit checklist

Before you leave, verify:

1. one improved plot
2. one merged pull request
3. one working Makefile rule
4. one explanation of the missing-prerequisite problem

---

## Solutions for core exercises

Use only if blocked.

---

## Solution - Exercise 1

```{r}
ggplot(videos, aes(x = impressions_n, y = watch_rate)) +
  geom_point()
```

---

## Solution - Exercise 2

```{r}
videos_plot <- videos %>%
  mutate(
    visibility_band = case_when(
      impressions_n < 20 ~ "Low visibility",
      impressions_n < 60 ~ "Medium visibility",
      TRUE ~ "High visibility"
    )
  )

ggplot(videos_plot, aes(impressions_n, watch_rate, color = visibility_band)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Video quality vs reach",
    x = "Impressions",
    y = "Watch rate",
    color = "Visibility"
  ) +
  theme_minimal()
```

---

## Solution - Exercise 2 reflection example

Possible reflection:

- Used: clearer legend title
- Not used: faceting by creator (too many panels)

---

## Solution - Exercise 3

```bash
git pull origin main
git switch -c feature-week3-readme
# edit one line in README.md
git status
git add README.md
git commit -m "Update README for week 3"
git push -u origin feature-week3-readme
```

---

## Solution - Exercise 4

On GitHub:

1. Open pull request from `feature-week3-readme` to `main`
2. Check files and diff
3. Merge pull request
4. Delete branch

---

## Solution - Exercise 5

`build_plot.R` and `makefile`:

```r
library(tidyverse)

url <- paste0(
  "https://raw.githubusercontent.com/hannesdatta/course-dprep/",
  "refs/heads/main/material/tutorials/r-bootcamp-rev/video_view.csv"
)
if (!file.exists("video_view.csv")) download.file(url, "video_view.csv")

videos <- read_csv("video_view.csv")

ggplot(videos, aes(impressions_n, watch_rate)) +
  geom_point(color = "steelblue", alpha = 0.7)

ggsave("plot_week3.png", width = 7, height = 4)
```

```make
plot_week3.png: build_plot.R video_view.csv
	Rscript build_plot.R
```

---

## Solution - Exercise 6

Correct rule:

```make
plot_week3.png: build_plot.R video_view.csv
	Rscript build_plot.R
```

Wrong rule:

```make
plot_week3.png: video_view.csv
	Rscript build_plot.R
```

Why wrong: edits in `build_plot.R` are ignored.

---

## Extra practice (optional)

Choose extra exercises if you finish early.

---

## Extra plots - Exercise P1

Create a histogram of `watch_rate`.

Add:

- clear title
- x and y labels
- `bins = 20`

---

## Extra plots - Exercise P2

Create a bar chart with top 10 creators by average watch rate.

Steps:

1. `group_by(creator_id)`
2. `summarise(avg_watch = mean(watch_rate, na.rm = TRUE))`
3. sort descending and keep 10
4. plot with `geom_col()`

---

## Extra plots - Exercise P3

Create a line chart of number of videos by publication date.

Hint:

- convert `published_at` with `as.POSIXct(..., origin = "1970-01-01")`
- create a date column
- count videos per date
- use `geom_line()`

---

## Solution - Extra plot P1

```{r}
ggplot(videos, aes(x = watch_rate)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of watch rate",
    x = "Watch rate",
    y = "Count"
  )
```

---

## Solution - Extra plot P2

```{r}
top_creators <- videos %>%
  group_by(creator_id) %>%
  summarise(avg_watch = mean(watch_rate, na.rm = TRUE)) %>%
  arrange(desc(avg_watch)) %>%
  slice_head(n = 10)

ggplot(top_creators, aes(x = reorder(as.factor(creator_id), avg_watch), y = avg_watch)) +
  geom_col(fill = "tomato") +
  coord_flip() +
  labs(
    title = "Top 10 creators by average watch rate",
    x = "Creator",
    y = "Average watch rate"
  )
```

---

## Solution - Extra plot P3

```{r}
daily_counts <- videos %>%
  mutate(pub_datetime = as.POSIXct(published_at, origin = "1970-01-01", tz = "UTC")) %>%
  mutate(pub_date = as.Date(pub_datetime)) %>%
  count(pub_date)

ggplot(daily_counts, aes(x = pub_date, y = n)) +
  geom_line(color = "darkgreen") +
  labs(
    title = "Videos published per day",
    x = "Publication date",
    y = "Number of videos"
  )
```

---

## Extra branching - Exercise B1

Run a second branch cycle with another branch name.

Use `README.md` again:

1. create branch `feature-week3-b2`
2. edit one line
3. `status -> add -> commit -> push`
4. open and merge PR

---

## Extra branching - Exercise B2

Inspect branch history after merges:

```bash
git log --oneline --graph --decorate -10
```

Question:

- Can you identify your merge commit?

---

## Solution - Extra branching B1

```bash
git switch main
git pull origin main
git switch -c feature-week3-b2
# edit README.md
git status
git add README.md
git commit -m "Add second branch exercise line"
git push -u origin feature-week3-b2
```

Then open and merge PR on GitHub.

---

## Solution - Extra branching B2

```bash
git log --oneline --graph --decorate -10
```

Look for a commit message like `Merge pull request ...`.

---

## Extra Make - Exercise M1

Add a second target that creates `summary_week3.csv`.

Create `summary.R`:

```r
library(tidyverse)
videos <- read_csv("video_view.csv")

summary_tbl <- videos %>%
  summarise(
    n_videos = n(),
    avg_watch = mean(watch_rate, na.rm = TRUE)
  )

write_csv(summary_tbl, "summary_week3.csv")
```

Then add this rule to your makefile:

```make
summary_week3.csv: summary.R video_view.csv
	Rscript summary.R
```

---

## Extra Make - Exercise M2

Add two convenience targets:

```make
all: plot_week3.png summary_week3.csv

wipe:
	rm -f plot_week3.png summary_week3.csv
```

Run:

```bash
make -n
make
make wipe
```

---

## Solutions - Extra Make

M1 rule:

```make
summary_week3.csv: summary.R video_view.csv
	Rscript summary.R
```

M2 rules:

```make
all: plot_week3.png summary_week3.csv

wipe:
	rm -f plot_week3.png summary_week3.csv
```

---

## End of tutorial

If you finish early, pick one extra exercise from each block:

1. one extra plot exercise
2. one extra branching exercise
3. one extra Make exercise
