---
title: "Tutorial (week 2) - Quarto, dplyr, Git, and automation"
subtitle: "Data Preparation and Programming Skills (2026/2027)"
author: "Javier Torralba Flores"
date: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
  message: false
format:
  beamer:
    theme: metropolis
    colortheme: seahorse
    code-overflow: wrap
    code-font-size: 0.9em
    aspectratio: 169
    fontsize: 11pt
    highlight-style: tango
    include-in-header:
      text: |
        \usepackage{xcolor}
        \newcommand{\workbg}{\setbeamercolor{background canvas}{bg=yellow!12}}
        \newcommand{\normalbg}{\setbeamercolor{background canvas}{bg=white}}
---

## Tutorial roadmap (90 minutes)

Today builds on last week with the same data file: `video_view.csv`.

1. Quarto + Markdown basics (15 min)
2. Coding with `dplyr` + aggregation (35 min)
3. Git collaboration workflow (20 min)
4. Automation with README + rerun commands (10 min)
5. Wrap-up + next steps (10 min)

You get dedicated work blocks with clear exercise numbers.

---

## Last week in 60 seconds

You already practiced this loop:

- Coding: `read_csv()`, inspect data, basic transformations
- Collaboration: `git status` -> `git add` -> `git commit` -> `git push`
- Automation: run scripts from terminal with `Rscript`

Today we keep the same loop, but make it cleaner and more reproducible.

---

## Thinking prompt (1 minute)

Discuss with your neighbor:

1. What was easy last week?
2. What was still confusing?
3. If you open your files in 3 weeks, what will help you understand your own work?

Some nudges: Markdown, readmes, dplyr, git, Rscript...

---

## Section 1: Quarto and Markdown

Goal: understand what Markdown is and how Quarto turns it into output.

---

## What is Markdown?

Markdown is plain text with lightweight formatting.

- `# Heading 1`
- `## Heading 2`
- `- bullet list`
- `[link text](https://example.com)`

Why this is useful: simple, readable, and version-friendly.

---

## What is Quarto?

Quarto is a rendering engine for `.qmd` files.

- you write text + code in one file
- you render it into HTML/PDF/slides
- your analysis and explanation stay together

Think: one source of truth for both code and communication.

---

## Minimal `.qmd` structure

````markdown
---
title: "My first Quarto page"
format: html
---

# Question

What does `video_view.csv` look like?

```{r}
library(tidyverse)
```
````

Three layers: YAML, narrative text, code chunks.

---

## Render command

```bash
quarto render report.qmd
```

Interactive check:

- Before running: what output file do you expect?
- After running: did that file appear in your folder?

---

## WORK TIME - Exercise 1 (8 minutes)

```{=latex}
\workbg
```

Create `report.qmd` that includes:

1. a title in YAML
2. one heading + two sentences in Markdown
3. one R code chunk with `library(tidyverse)`
4. one R code chunk with `getwd()` and `list.files()`
5. render to HTML

Expected output checklist:

- `report.html` exists
- HTML shows both your text and code output

---

## Section 2: Coding with `video_view.csv`

```{=latex}
\normalbg
```

Goal: practice core `dplyr` verbs and aggregation.

No merges/joins yet.

---

## Load data and inspect

```{r}
library(tidyverse)

if (!file.exists("video_view.csv")) {
  data_url <- paste0(
    "https://raw.githubusercontent.com/hannesdatta/",
    "course-dprep/refs/heads/main/material/tutorials/r-bootcamp-rev/",
    "video_view.csv"
  )
  download.file(data_url, "video_view.csv")
}

videos <- read_csv("video_view.csv")
```

Inspect:

```{r}
head(videos)
names(videos)
glimpse(videos)
summary(videos$watch_rate)
```

---

## Think before coding

Look at the column names.

Question to discuss:

- Which columns seem useful for creator performance?
- Which columns might help compare quality vs reach?

Then test your ideas with code.

---

## Core `dplyr` verbs this week

- `select()` keep columns
- `filter()` keep rows
- `mutate()` create new variables
- `rename()` rename variables
- `arrange()` sort rows
- `group_by()` define groups
- `summarise()` aggregate within groups

---

## Guided example 1: select + filter (question + plan)

Question:

- Which videos are both high quality and reasonably visible?

Predict first:

- If we require `watch_rate >= 0.7` and `impressions_n >= 20`, do we keep many or few rows?

Plan:

1. keep only needed columns (`select`)
2. keep only rows that pass both conditions (`filter`)

```{r}
high_quality <- videos %>%
  select(video_id, creator_id, impressions_n, watched_n, watch_rate) %>%
  filter(watch_rate >= 0.7, impressions_n >= 20)
```

---

## Guided example 1: line-by-line explanation

```{r}
high_quality <- videos %>%
  select(video_id, creator_id, impressions_n, watched_n, watch_rate) %>%
  filter(watch_rate >= 0.7, impressions_n >= 20)
```

What each line does:

- `videos %>%` starts from the full table
- `select(...)` keeps only 5 relevant variables
- `filter(...)` keeps rows passing both rules

Check your result:

- run `nrow(high_quality)` and compare with `nrow(videos)`

Common beginner mistake:

- using `watch_rate = 0.7` instead of `watch_rate >= 0.7`

---

## Guided example 2: mutate + rename + arrange (question + plan)

Question:

- How can we create and rank a simple video performance score?

Predict first:

- Will `watched_n / impressions_n` likely be close to `watch_rate`?

Plan:

1. create new variable `watched_share`
2. rename long variable names for readability
3. sort descending to see top rows first

```{r}
video_scores <- videos %>%
  mutate(
    watched_share = watched_n / impressions_n
  ) %>%
  rename(
    impressions = impressions_n,
    watched = watched_n
  ) %>%
  arrange(desc(watched_share))
```

---

## Guided example 2: line-by-line explanation

```{r}
video_scores <- videos %>%
  mutate(
    watched_share = watched_n / impressions_n
  ) %>%
  rename(
    impressions = impressions_n,
    watched = watched_n
  ) %>%
  arrange(desc(watched_share))
```

What each line does:

- `mutate(...)` adds a new column without deleting old ones
- `rename(...)` makes names shorter and easier to read
- `arrange(desc(...))` puts largest values at the top

Check your result:

- run `video_scores %>% select(video_id, watched_share) %>% head()`

Common beginner mistake:

- forgetting `desc()` and accidentally sorting smallest-to-largest

---

## Guided example 3: aggregation without groups

Question:

- What are overall totals/averages in the full dataset?

```{r}
videos %>%
  summarise(
    total_impressions = sum(impressions_n, na.rm = TRUE),
    avg_watch_rate = mean(watch_rate, na.rm = TRUE),
    n_videos = n()
  )
```

What this returns:

- exactly one row with full-sample metrics

Check:

- confirm with `nrow(...)` that output has one row

---

## Guided example 4: group_by + summarise

Question:

- How do creator-level metrics differ across creators?

```{r}
creator_summary <- videos %>%
  group_by(creator_id) %>%
  summarise(
    videos_n = n(),
    impressions_total = sum(impressions_n, na.rm = TRUE),
    watch_rate_avg = mean(watch_rate, na.rm = TRUE)
  ) %>%
  arrange(desc(impressions_total))
```

What this returns:

- one row per `creator_id`

Check:

- `n_distinct(videos$creator_id)` should match `nrow(creator_summary)`

Common beginner mistake:

- forgetting `group_by(creator_id)` and getting only one summary row

---

## WORK TIME - Exercise 2 (10 minutes)

```{=latex}
\workbg
```

Create `creator_summary` from `videos` with:

1. `group_by(creator_id)`
2. `summarise(videos_n, impressions_total, watch_rate_avg)`
3. sorted by `impressions_total` descending

Then answer in one sentence:

- Which creator has the most impressions?

Starter code:

```{r}
creator_summary <- videos %>%
  group_by(creator_id) %>%
  summarise(
    videos_n = ___,
    impressions_total = ___,
    watch_rate_avg = ___
  ) %>%
  arrange(desc(___))
```

---

## WORK TIME - Exercise 3 (8 minutes)

```{=latex}
\workbg
```

Create object `top_watch` with:

1. `select(video_id, creator_id, impressions_n, watch_rate)`
2. `filter(impressions_n >= 50)`
3. `arrange(desc(watch_rate))`

Then answer:

- Is the top-ranked video also high in impressions?

Starter code:

```{r}
top_watch <- videos %>%
  select(___) %>%
  filter(___) %>%
  arrange(___)
```

---

## WORK TIME - Exercise 4 (8 minutes)

```{=latex}
\workbg
```

Add a new variable and summarize it:

1. create `engagement = watched_n / impressions_n` using `mutate()`
2. aggregate by `creator_id` using `group_by()` + `summarise()`
3. compute mean engagement per creator
4. sort descending

Expected output: one row per creator with mean engagement.

Starter code:

```{r}
creator_engagement <- videos %>%
  mutate(engagement = ___ / ___) %>%
  group_by(___) %>%
  summarise(mean_engagement = mean(___, na.rm = TRUE)) %>%
  arrange(desc(___))
```

---

## Section 3: Collaboration with Git

```{=latex}
\normalbg
```

Goal: repeat the commit cycle and add clone/pull/fork practice.

---

## Collaboration workflow (beginner-safe)

```bash
git clone <repo-url>
git pull origin main
git status
git add report.qmd
git commit -m "Add creator summary exercises"
git push origin main
```

Interactive prompt:

- Which command tells you what changed locally?
- Which command sends your local commits to GitHub?

---

## Fork workflow for class practice

Practice sequence:

1. Fork classmate repository on GitHub
2. Clone your fork locally
3. Pull before editing
4. Make one small README improvement
5. Run `status` -> `add` -> `commit` -> `push`

No merge conflict handling yet.

---

## WORK TIME - Exercise 5 (10 minutes, pairs)

```{=latex}
\workbg
```

Pair task:

1. Student A forks Student B repo
2. Student A clones fork
3. Student A updates one README line
4. Student A commits and pushes
5. Student B verifies update on GitHub

Expected output checklist:

- a new commit visible on Student A fork
- Student B can see the change online

---

## Section 4: Automation for beginners

```{=latex}
\normalbg
```

Goal: make your project easy to rerun for yourself and others.

---

## Why README matters

README is lightweight automation for humans.

It should answer:

1. What is this project?
2. Which data file is needed?
3. Which commands should I run, and in what order?
4. What outputs should I expect?

---

## README template

```markdown
# Project title

## Goal
Short project objective.

## Data
- Main file: video_view.csv

## Requirements
- R
- Packages: tidyverse
- Quarto installed

## Run steps
1. `Rscript analysis.R`
2. `quarto render report.qmd`

## Expected output
- report.html
```

---

## WORK TIME - Exercise 6 (8 minutes)

```{=latex}
\workbg
```

Create or update `README.md` with:

1. goal
2. data (`video_view.csv`)
3. requirements
4. exact run commands
5. expected output

Then commit your README changes.

---

## Wrap-up and where to go next

```{=latex}
\normalbg
```

You now have a complete beginner workflow:

- Quarto document with Markdown + code
- `dplyr` verbs and aggregation with one dataset
- Git sync and commit cycle
- README-driven rerun instructions

Next: deeper wrangling and then table-combining later (not yet).

---

## Appendix A: extra practice questions

1. Count videos with `watch_rate >= 0.8`.
2. Find the top 5 creators by average `watch_rate`.
3. Find creators with at least 3 videos and sort by total impressions.
4. Create a short interpretation paragraph (3 lines) for your `creator_summary`.
5. Add a section to README called "Troubleshooting" with 3 common errors.
6. Render your Quarto file again after one small text edit and verify output updates.

---

## Solutions section starts here

All solutions are visible and matched by exercise number.

---

## Solution - Exercise 1

Example `report.qmd`:

````markdown
---
title: "Week 2 practice"
format: html
---

# First report

This is my first Quarto practice file.
I combine Markdown text and R code.

```{r}
library(tidyverse)
```

```{r}
getwd()
list.files()
```
````

Render command:

```bash
quarto render report.qmd
```

---

## Solution - Exercise 2

```{r}
creator_summary <- videos %>%
  group_by(creator_id) %>%
  summarise(
    videos_n = n(),
    impressions_total = sum(impressions_n, na.rm = TRUE),
    watch_rate_avg = mean(watch_rate, na.rm = TRUE)
  ) %>%
  arrange(desc(impressions_total))

creator_summary
```

Interpretation example: "`creator_id` at the top has the highest total impressions in this dataset."

---

## Solution - Exercise 3

```{r}
top_watch <- videos %>%
  select(video_id, creator_id, impressions_n, watch_rate) %>%
  filter(impressions_n >= 50) %>%
  arrange(desc(watch_rate))

top_watch
```

Interpretation example: "The top watch-rate video may or may not have the most impressions; quality and reach can differ."

---

## Solution - Exercise 4

```{r}
creator_engagement <- videos %>%
  mutate(engagement = watched_n / impressions_n) %>%
  group_by(creator_id) %>%
  summarise(mean_engagement = mean(engagement, na.rm = TRUE)) %>%
  arrange(desc(mean_engagement))

creator_engagement
```

Interpretation example: "This ranks creators by average engagement instead of raw reach."

---

## Solution - Exercise 5

Commands used in the workflow:

```bash
git clone <fork-url>
git pull origin main
git status
git add README.md
git commit -m "Improve README instructions"
git push origin main
```

Success check:

- commit is visible on GitHub fork
- changed README content appears online

---

## Solution - Exercise 6

Example `README.md`:

```markdown
# Video analysis practice

## Goal
Practice Quarto, dplyr aggregation, and Git workflow.

## Data
- video_view.csv

## Requirements
- R
- tidyverse package
- Quarto

## Run steps
1. Rscript analysis.R
2. quarto render report.qmd

## Expected output
- report.html

## Troubleshooting
- If file not found: run getwd() and list.files().
- If tidyverse missing: install.packages("tidyverse").
- If render fails: check Quarto installation.
```

---

## End of tutorial

If you finish early, continue with Appendix A extra practice.
