---
title: "Tutorial (week 4) - Data engineering, pull requests, and Make"
subtitle: "Data Preparation and Programming Skills (2026/2027)"
author: "Javier Torralba Flores"
date: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
  message: false
format:
  beamer:
    theme: metropolis
    colortheme: seahorse
    code-overflow: wrap
    code-font-size: 0.85em
    aspectratio: 169
    fontsize: 11pt
    highlight-style: tango
    include-in-header:
      text: |
        \usepackage{xcolor}
        \newcommand{\workbg}{\setbeamercolor{background canvas}{bg=yellow!12}}
        \newcommand{\normalbg}{\setbeamercolor{background canvas}{bg=white}}
---

## Tutorial roadmap (90 minutes)

Today follows one project story with four blocks:

1. Opening and orientation - 10 min
2. Coding and frequent commits in branch A - 50 min
3. Automation and frequent commits in branch B - 20 min
4. Pull requests: merge PR1 and PR2 - 10 min

Total: 90 minutes.

---

## Opening story

Imagine this week in your team:

- Monday: you build creator and user metrics from raw CSV files.
- Tuesday: you commit small coding changes in a feature branch.
- Wednesday: a teammate builds Make automation in another branch.
- Thursday: both branches are reviewed in pull requests.
- Friday: both PRs are merged into `main` and outputs rebuild cleanly.

Today you practice this exact flow.

---

## Today’s flow

You will produce two pull requests:

1. **PR1** = `feature/week4-data` -> `main` (coding outputs)
2. **PR2** = `feature/week4-make` -> `main` (automation outputs)

Each PR should contain multiple small commits, not one giant commit.

---

## Git + PR quick refresher

- `git status`: what changed in your working tree?
- `git add <file>`: stage file(s) for the next commit.
- `git commit -m "..."`: save a labeled snapshot.
- `git merge <branch>`: combine another branch into current branch.
- Pull request (PR): GitHub proposal to merge one branch into another.

In this class, merges happen through GitHub PRs.

---

## Pull vs pull request

- `git pull` = receive updates from remote repository.
- pull request = propose your updates for review and merge.

Short memory trick:

- pull = download
- pull request = propose

---

## Learning goals

By the end, you can:

- create variables with `mutate()`, `rank()`, and `distinct()`
- aggregate metrics with `group_by()` and `summarise()`
- combine tables with `left_join()` and `inner_join()`
- convert timestamps and build lagged metrics
- work with two branches, frequent commits, and two pull requests
- automate outputs with `make` and `make clean`

---

## Coding block

```{=latex}
\normalbg
```

Today’s coding questions:

- Which creators have both high reach and high watch quality?
- Which users show strong engagement?
- How does creator exposure change day by day?
- Can a teammate read your code and merge PR1 confidently?

---

## Download the data if you don't have it yet 

```{r}
library(tidyverse)

base_url <- paste0(
  "https://raw.githubusercontent.com/hannesdatta/",
  "course-dprep/refs/heads/main/material/project/"
)

files <- c("video_view.csv",
           "coaching_2_data/impressions.csv",
           "coaching_2_data/watch_events.csv")

for (f in files) {
  dest <- basename(f)
  if (!file.exists(dest)) download.file(paste0(base_url, f), dest)}
```

---

## Data preview (video-level)

Run a quick peek so the tasks feel concrete:

```{r}
video_view <- read_csv("video_view.csv")

video_view %>%
  select(video_id, creator_id, impressions_n, watched_n, watch_rate, avg_watch_share) %>%
  head(6)
```

---

## Data preview (event-level)

```{r}
impressions <- read_csv("impressions.csv")
watch_events <- read_csv("watch_events.csv")

impressions %>%
  select(impression_id, session_id, user_id, video_id, creator_id, shown_at) %>%
  head(6)
```

---

## Start here

1. Create/open your week 4 project.
2. Put all provided CSV files in your project root.
3. Create folders for scripts and generated outputs.

Main script for this block:

`src/data_engineering.R`

Commit rule for today:

- commit scripts and makefiles
- do not commit generated CSV files from `temp/` or `output/`

---

## Create folders (three options)

Pick one method:

- Positron/VS Code file pane: create `src`, `temp`, `output` manually.
- macOS/Linux terminal:

```bash
mkdir -p src temp output
```

- Windows PowerShell:

```powershell
mkdir src, temp, output
```

---

## Why branch A first?

`main` should stay stable.

- Branch A is your safe space for data-engineering work.
- You can make many commits without touching `main`.
- PR1 is the place where others review your coding decisions.

---

## Start branch A (CLI)

```bash
git pull origin main
git switch -c feature/week4-data
git branch --show-current
git status
```

If successful, you should see:

- current branch: `feature/week4-data`
- `On branch feature/week4-data` in status output

---

## Start branch A (Positron/VS Code)

Use GUI if preferred:

1. Open **Source Control** panel.
2. Click branch name in status bar.
3. Select **Create new branch** from `main`.
4. Name it `feature/week4-data`.
5. Confirm active branch in status bar.

Still run `git status` in terminal to verify.

---

## Load data

Add this at the top of `src/data_engineering.R`:

```{r}
library(tidyverse)

video_view <- read_csv("video_view.csv")
user_view <- read_csv("user_view.csv")
videos <- read_csv("videos.csv")
creators <- read_csv("creators.csv")
users <- read_csv("users.csv")
impressions <- read_csv("impressions.csv")
watch_events <- read_csv("watch_events.csv")
sessions <- read_csv("sessions.csv")
```

---

## Function focus for this block

In order:

1. `mutate()`, `rank()`, `distinct()`
2. `group_by()`, `summarise()`, `arrange()`
3. `left_join()`, `inner_join()`
4. `as.POSIXct()`, `as.Date()`, `lag()`
5. `write_csv()` for clear output files (ITO)

---

## Example: start simple with rank

```{r}
video_simple <- video_view %>%
  mutate(
    watch_rate_rank = rank(-watch_rate),
    high_quality = avg_watch_share >= 0.40
  ) %>%
  distinct(video_id, .keep_all = TRUE)
```

---

## Optional rank settings (later)

After the simple version works, you can refine ties and missing values:

```{r}
video_ranked <- video_view %>%
  mutate(
    watch_rate_rank = rank(-watch_rate, na.last = "keep", ties.method = "min")
  )
```

Use this only after understanding the simple rank first.

---

## WORK TIME - Exercise 1

```{=latex}
\workbg
```

Story: you prepare the first PR1 output table.

In `src/data_engineering.R`, build `video_features`:

1. start from `video_view`
2. create `watch_rate_rank = rank(-watch_rate)`
3. create `reach_band` (`Low`, `Medium`, `High` from `impressions_n`)
4. create `high_quality` (`avg_watch_share >= 0.40`)
5. keep unique `video_id` rows with `distinct()`
6. save output: `write_csv(video_features, "temp/video_features.csv")`

Show top 10 rows sorted by `watch_rate_rank`.

---

## Git checkpoint A1

```{=latex}
\workbg
```

Run and commit:

```bash
git status
git add src/data_engineering.R
git commit -m "Add video features with mutate rank and distinct"
git status
```

---

## What is aggregation?

Aggregation means:

- go from many rows -> fewer summary rows
- summarize by groups (for example by creator)

Example idea:

- 1200 video rows -> 60 creator rows

This is one of the most common operations in analytics.

---

## Example: aggregation

```{r}
creator_example <- video_view %>%
  group_by(creator_id) %>%
  summarise(
    impressions_total = sum(impressions_n, na.rm = TRUE),
    avg_watch_rate = mean(watch_rate, na.rm = TRUE)
  ) %>%
  arrange(desc(impressions_total))
```

---

## WORK TIME - Exercise 2

```{=latex}
\workbg
```

Create two aggregate tables from `video_features`:

1. `creator_summary` with:
   - `videos_n`
   - `impressions_total`
   - `watched_total`
   - `avg_watch_rate`
   - `median_watch_seconds`
2. `engagement_by_band` with:
   - number of videos
   - average watch rate

Sort `creator_summary` by `desc(impressions_total)`.

---

## Exercise 2 output (ITO)

```{=latex}
\workbg
```

Save both outputs:

```r
write_csv(creator_summary, "temp/creator_summary.csv")
write_csv(engagement_by_band, "temp/engagement_by_band.csv")
```

---

## Git checkpoint A2

```{=latex}
\workbg
```

```bash
git status
git add src/data_engineering.R
git commit -m "Add creator and reach-band aggregations"
git status
```

---

## Join theory: why joins matter

Our data is spread across files.

- behavior: `video_view.csv`, `user_view.csv`, `watch_events.csv`
- context: `creators.csv`, `videos.csv`, `users.csv`, `sessions.csv`

Joins combine them into analysis-ready tables.

---

## Join theory: two core join types

- `left_join(x, y, by = ...)`
  - keep all rows from `x`
  - add columns from `y` when keys match
- `inner_join(x, y, by = ...)`
  - keep only matched rows in both tables

Default in data prep: start with `left_join()`.

---

## Join theory: keys and risks

- A key is the variable used to match rows.
- Single key example: `creator_id`
- Composite key example: `video_id` + `creator_id`

Risk:

- repeated keys on both sides can duplicate rows.

Always check row counts before and after big joins.

---

## Example: one left join and one inner join

```{r}
video_with_creators <- video_view %>%
  left_join(creators, by = "creator_id")

watched_only <- impressions %>%
  inner_join(watch_events, by = "impression_id")
```

---

## WORK TIME - Exercise 3

```{=latex}
\workbg
```

Build `video_enriched`:

1. start from `video_features`
2. `left_join(videos, by = c("video_id", "creator_id"))`
3. `left_join(creators, by = "creator_id")`
4. keep columns:
   - `video_id`, `creator_id`, `creator_name`
   - `impressions_n`, `watch_rate`, `watch_rate_rank`
   - `quality`, `posting_rate`, `publish_time`

---

## WORK TIME - Exercise 3 (continued)

```{=latex}
\workbg
```

Build `user_enriched`:

1. start from `user_view`
2. `left_join(users, by = "user_id")`
3. keep user watch KPIs and selected user profile columns
4. save outputs:
   - `temp/video_enriched.csv`
   - `temp/user_enriched.csv`

---

## WORK TIME - Exercise 4

```{=latex}
\workbg
```

Build event-level `watch_log` in steps:

1. start from `impressions`
2. `left_join(watch_events, by = "impression_id")`
3. `left_join(sessions, by = c("session_id", "user_id"))`
4. `left_join(videos, by = c("video_id", "creator_id"))`
5. `left_join(creators, by = "creator_id")`

---

## WORK TIME - Exercise 4 (continued)

```{=latex}
\workbg
```

From the same data, create:

1. `watched_only <- impressions %>% inner_join(watch_events, by = "impression_id")`
2. `creator_event_summary` with:
   - impressions
   - watched events
   - total watch seconds

Save `watch_log` and `creator_event_summary` in `temp/`.

---

## Git checkpoint A3

```{=latex}
\workbg
```

```bash
git status
git add src/data_engineering.R
git commit -m "Add join workflows for view and event data"
git status
```

---

## Timestamp conversion first

Focus on conversion only:

- `shown_at` is text in ISO format (example: `2025-08-01T21:44:48Z`)
- convert text -> datetime (`shown_ts`, class `POSIXct`)
- convert datetime -> day (`shown_day`, class `Date`)

Why this conversion?

- datetime is needed for time-aware operations
- day is needed for daily aggregation per creator

```{r}
watch_time_preview <- watch_log %>%
  mutate(
    shown_ts = as.POSIXct(shown_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    shown_day = as.Date(shown_ts)
  ) %>%
  select(impression_id, creator_id, shown_at, shown_ts, shown_day) %>%
  head(8)
```

---

## What the daily lag table creates

Once timestamp conversion works:

- one row per `creator_id` and `shown_day`
- `impressions_n`: number of impressions on that day
- `impressions_lag1`: previous day's impressions for that creator

Then you can compute:

- `impressions_change = impressions_n - impressions_lag1`

```{r}
creator_daily <- watch_log %>%
  mutate(shown_ts = as.POSIXct(shown_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")) %>%
  mutate(shown_day = as.Date(shown_ts)) %>%
  count(creator_id, shown_day, name = "impressions_n") %>%
  group_by(creator_id) %>%
  arrange(shown_day) %>%
  mutate(impressions_lag1 = lag(impressions_n))
```

---

## WORK TIME - Exercise 5

```{=latex}
\workbg
```

From `watch_log`:

1. convert `shown_at` into `shown_ts` and `shown_day`
2. aggregate daily creator impressions
3. add:
   - `impressions_lag1 = lag(impressions_n)`
   - `impressions_change = impressions_n - impressions_lag1`
4. save `temp/creator_daily_week4.csv`

---

## Git checkpoint A4

```{=latex}
\workbg
```

```bash
git status
git add src/data_engineering.R
git commit -m "Add timestamp conversion and lagged daily metrics"
git status
```

---

## Inspect your branch history

Before switching to branch B, inspect what you built:

```bash
git status
git log --oneline --graph -10
git diff HEAD~1 HEAD
```

Questions:

- Do your commit messages tell a clear story?
- Can you see exactly what changed in the last commit?

---

## Push branch A

```bash
git push -u origin feature/week4-data
```

This branch will become **PR1** later.

---

## Automation block

```{=latex}
\normalbg
```

Now build a second branch for automation only.

Goal: reproducible outputs and clean rebuilds.

---

## Why branch B?

- Branch A holds data-engineering code for PR1.
- Branch B isolates Make changes for PR2.
- Separate PRs make review easier.

This is common in team workflows.

---

## Start branch B

```bash
git switch main
git pull origin main
git switch -c feature/week4-make
git branch --show-current
git status
```

Expected branch: `feature/week4-make`.

---

## Switch branch in Positron/VS Code

GUI route:

1. Open **Source Control**.
2. Click branch name in the bottom bar.
3. Choose `main`, then create `feature/week4-make`.
4. Confirm active branch in the bottom bar.

Check with `git status` after switching.

---

## Make theory: why we need it

Running scripts manually is easy to forget.

`make` gives a recipe so the computer can:

- run steps in the right order
- rerun only what changed
- rebuild from scratch when needed

Think cookbook for code.

---

## Make theory: one rule

A rule has three parts:

```text
target: prerequisites
<TAB>command to build target
```

- target: file to create
- prerequisites: files needed first
- command: how to build

Commands must start with a **tab**, not spaces.

---

## Make theory: dependencies and dry run

- If a prerequisite changes, `make` rebuilds dependent targets.
- `make -n` shows what would run, without running it.

Useful workflow:

1. `make -n`
2. check output
3. run `make`

---

## Make theory: all and clean

- `all` is usually the first target and default entry point.
- `clean` removes generated outputs.

Common pattern:

```text
all: final_output
clean:
<TAB>remove generated files
```

---

## Script structure for automation

Use Setup-Input-Transform-Output in each script.

Files for this block:

- `src/build_temp.R`
- `src/build_output.R`
- `makefile`

---

## WORK TIME - Exercise 6

```{=latex}
\workbg
```

Create `src/build_temp.R`:

1. input `video_view.csv` and `creators.csv`
2. aggregate to creator level
3. output `temp/creator_week4.csv`

Create `src/build_output.R`:

1. input `temp/creator_week4.csv`
2. keep top 10 creators by impressions
3. output `output/creator_top10_week4.csv`

---

## Git checkpoint B1

```{=latex}
\workbg
```

```bash
git status
git add src/build_temp.R src/build_output.R
git commit -m "Add build scripts for temp and output tables"
git status
```

---

## WORK TIME - Exercise 7

```{=latex}
\workbg
```

Create this `makefile`:

```make
all: output/creator_top10_week4.csv

temp/creator_week4.csv: src/build_temp.R video_view.csv creators.csv
	Rscript src/build_temp.R

output/creator_top10_week4.csv: src/build_output.R temp/creator_week4.csv
	Rscript src/build_output.R

clean:
	rm -f temp/*.csv output/*.csv
```

---

## Exercise 7 run commands

```{=latex}
\workbg
```

```bash
make -n
make
make
make clean
make
```

What to observe:

- first run builds files
- second run should do almost nothing
- `make clean` removes generated files
- final `make` rebuilds all targets

---

## Git checkpoint B2

```{=latex}
\workbg
```

```bash
git status
git add makefile
git commit -m "Add make workflow with clean target"
git status
git log --oneline --graph -10
git diff HEAD~1 HEAD
git push -u origin feature/week4-make
```

This branch will become **PR2** later.

---

## PR block

```{=latex}
\normalbg
```

Now merge both branches through GitHub pull requests.

---

## Where pull requests are created

PRs are created on **GitHub.com**, not in local files.

Path:

1. Open your repository on GitHub.
2. Click **Pull requests** tab.
3. Click **New pull request**.
4. Choose base `main` and compare branch.

---

## PR1 checklist

PR1 uses coding branch:

- compare: `feature/week4-data`
- base: `main`
- title example: `Week 4 data engineering outputs`
- confirm commit list and changed files
- request review (if working in team)

Merge PR1 when ready.

---

## PR2 checklist

PR2 uses automation branch:

- compare: `feature/week4-make`
- base: `main`
- title example: `Week 4 make automation and clean target`
- confirm commit list and changed files
- request review

Merge PR2 when ready.

---

## WORK TIME - Exercise 8

```{=latex}
\workbg
```

Do both PR merges on GitHub:

1. open PR1 and merge
2. open PR2 and merge
3. delete both remote branches

After merges, sync local repository.

---

## Local cleanup after PR merges

```{=latex}
\workbg
```

```bash
git switch main
git pull origin main
git branch -d feature/week4-data
git branch -d feature/week4-make
git status
```

---

## Wrap-up

```{=latex}
\normalbg
```

You practiced one complete workflow:

- feature engineering and aggregation
- table joins for analysis-ready data
- timestamp conversion and lagging
- frequent commits in two branches
- make automation with clean rebuild
- two pull requests merged into `main`

---

## Exit checklist

Before you leave, verify:

1. PR1 and PR2 are both merged
2. `src/data_engineering.R` runs and writes outputs
3. `make clean` and `make` both work
4. your commit history tells a clear story

---

## Solutions for core exercises

Use only if blocked.

---

## Solution - Exercise 1 (build features)

```{r}
video_features <- video_view %>%
  mutate(
    watch_rate_rank = rank(-watch_rate),
    reach_band = case_when(
      impressions_n < 20 ~ "Low",
      impressions_n < 60 ~ "Medium",
      TRUE ~ "High"
    ),
    high_quality = avg_watch_share >= 0.40
  ) %>%
  distinct(video_id, .keep_all = TRUE) %>%
  arrange(watch_rate_rank)
```

---

## Solution - Exercise 1 (output)

```{r}
write_csv(video_features, "temp/video_features.csv")

video_features %>%
  select(video_id, creator_id, impressions_n, watch_rate, watch_rate_rank, reach_band, high_quality) %>%
  slice_head(n = 10)
```

---

## Solution - Exercise 2 (creator summary)

```{r}
creator_summary <- video_features %>%
  group_by(creator_id) %>%
  summarise(
    videos_n = n(),
    impressions_total = sum(impressions_n, na.rm = TRUE),
    watched_total = sum(watched_n, na.rm = TRUE),
    avg_watch_rate = mean(watch_rate, na.rm = TRUE),
    median_watch_seconds = median(total_watch_seconds, na.rm = TRUE)
  ) %>%
  arrange(desc(impressions_total))
```

---

## Solution - Exercise 2 (band summary + save)

```{r}
engagement_by_band <- video_features %>%
  group_by(reach_band) %>%
  summarise(
    videos_n = n(),
    avg_watch_rate = mean(watch_rate, na.rm = TRUE)
  )

write_csv(creator_summary, "temp/creator_summary.csv")
write_csv(engagement_by_band, "temp/engagement_by_band.csv")
```

---

## Solution - Exercise 3 (video join)

```{r}
video_enriched <- video_features %>%
  left_join(videos, by = c("video_id", "creator_id")) %>%
  left_join(creators, by = "creator_id") %>%
  select(
    video_id, creator_id, creator_name,
    impressions_n, watch_rate, watch_rate_rank,
    quality, posting_rate, publish_time
  )
```

---

## Solution - Exercise 3 (user join + save)

```{r}
user_enriched <- user_view %>%
  left_join(users, by = "user_id") %>%
  select(
    user_id, user_name, user_handle,
    impressions_n, watched_n, watch_rate,
    like_n, follow_n, baseline_login, satiation_decay
  )

write_csv(video_enriched, "temp/video_enriched.csv")
write_csv(user_enriched, "temp/user_enriched.csv")
```

---

## Solution - Exercise 4 (watch_log step 1)

```{r}
watch_log <- impressions %>%
  left_join(watch_events, by = "impression_id") %>%
  left_join(sessions, by = c("session_id", "user_id"))
```

---

## Solution - Exercise 4 (watch_log step 2)

```{r}
watch_log <- watch_log %>%
  left_join(videos, by = c("video_id", "creator_id")) %>%
  left_join(creators, by = "creator_id")

watched_only <- impressions %>%
  inner_join(watch_events, by = "impression_id")
```

---

## Solution - Exercise 4 (summary + save)

```{r}
creator_event_summary <- watch_log %>%
  group_by(creator_id) %>%
  summarise(
    impressions_n = n(),
    watched_events_n = sum(!is.na(action)),
    watch_seconds_total = sum(watch_seconds, na.rm = TRUE)
  )

write_csv(watch_log, "temp/watch_log.csv")
write_csv(creator_event_summary, "temp/creator_event_summary.csv")
```

---

## Solution - Exercise 5 (timestamp conversion)

```{r}
creator_daily <- watch_log %>%
  mutate(
    shown_ts = as.POSIXct(shown_at, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    shown_day = as.Date(shown_ts)
  ) %>%
  count(creator_id, shown_day, name = "impressions_n")
```

---

## Solution - Exercise 5 (lag + save)

```{r}
creator_daily <- creator_daily %>%
  group_by(creator_id) %>%
  arrange(shown_day) %>%
  mutate(
    impressions_lag1 = lag(impressions_n),
    impressions_change = impressions_n - impressions_lag1
  )

write_csv(creator_daily, "temp/creator_daily_week4.csv")
```

---

## Solution - Exercise 6 (`src/build_temp.R`)

```r
library(tidyverse)

dir.create("temp", showWarnings = FALSE)

video_view <- read_csv("video_view.csv")
creators <- read_csv("creators.csv")

creator_week4 <- video_view %>%
  group_by(creator_id) %>%
  summarise(
    videos_n = n(),
    impressions_total = sum(impressions_n, na.rm = TRUE),
    watched_total = sum(watched_n, na.rm = TRUE),
    avg_watch_rate = mean(watch_rate, na.rm = TRUE)
  ) %>%
  left_join(creators, by = "creator_id")
```

---

## Solution - Exercise 6 (`src/build_temp.R` output)

```r
write_csv(creator_week4, "temp/creator_week4.csv")
```

---

## Solution - Exercise 6 (`src/build_output.R`)

```r
library(tidyverse)

dir.create("output", showWarnings = FALSE)

creator_week4 <- read_csv("temp/creator_week4.csv")

creator_top10_week4 <- creator_week4 %>%
  arrange(desc(impressions_total)) %>%
  slice_head(n = 10)

write_csv(creator_top10_week4, "output/creator_top10_week4.csv")
```

---

## Solution - Exercise 7 (`makefile`)

```make
all: output/creator_top10_week4.csv

temp/creator_week4.csv: src/build_temp.R video_view.csv creators.csv
	Rscript src/build_temp.R

output/creator_top10_week4.csv: src/build_output.R temp/creator_week4.csv
	Rscript src/build_output.R

clean:
	rm -f temp/*.csv output/*.csv
```

---

## Solution - Exercise 7 (run)

```bash
make -n
make
make clean
make
```

---

## Solution - Exercise 8 (PRs)

On GitHub:

1. Open and merge PR1 from `feature/week4-data`.
2. Open and merge PR2 from `feature/week4-make`.
3. Delete both remote branches.

Then locally:

```bash
git switch main
git pull origin main
git branch -d feature/week4-data
git branch -d feature/week4-make
git status
```

---

## End of tutorial

If you finish early:

1. rerun `src/data_engineering.R` from top to bottom
2. run `make clean` and `make` again
3. inspect your Git story with `git log --oneline --graph -10`
