---
title: "Tutorial (week 6) - LLMs, prompting, and AI agents"
subtitle: "Data Preparation and Programming Skills (2026/2027)"
author: "Hannes Datta"
date: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
  message: false
format:
  beamer:
    theme: metropolis
    colortheme: seahorse
    code-overflow: wrap
    code-font-size: 0.85em
    aspectratio: 169
    fontsize: 11pt
    highlight-style: tango
    include-in-header:
      text: |
        \usepackage{xcolor}
        \newcommand{\workbg}{\setbeamercolor{background canvas}{bg=yellow!12}}
        \newcommand{\normalbg}{\setbeamercolor{background canvas}{bg=white}}
---

## Before we started

- [Code quests](https://code-quest.hannesdatta.com/) updated for all weeks - see https://code-quest.hannesdatta.com
- Exam prep. - example questions document updated for this year & posted on Canvas (11.45am today)
- Added [chapter 9](https://dprep-book.hannesdatta.com/ai-tooling-for-research) to the dPrep book at https://dprep-book.hannesdatta.com ("AI Tooling for Research: LLMs and Agentic Workflows")
- Set up for today[^1]
  - Download and unzip the **Week 6 starter project** from Canvas. The bundle includes the starter project and `data/tiktok_students.sqlite`.
  - Open the extracted `starter_repo` folder in RStudio or your preferred editor. Remember to set your working directories.

[^1]: Check the following files exist: `data/tiktok_students.sqlite`, `data/sample_video_view.csv`, `src/export_table.R`, `src/build_creator_summary.R`, `report.qmd`.

---

## Today's learning goals

By the end of this tutorial, you can:

- improve prompts by adding relevant course knowledge and checking model output;
- decide what the analyst must specify and what an LLM can help implement;
- use an agent to write and test an R script that exports SQLite views;
- use an agent to build and verify a Makefile that you can explain;
- (optional extra practice) use `lapply()` to repeat plots and regressions.

---

## LLMs: how a response is generated

- An LLM processes text as **tokens** and generates a response one token at a time.
- Its answer draws on patterns learned during training and the context provided now.
- It does not automatically know the contents of your files or the meaning of your variables.

It can explain code and draft useful responses. __Remember that its fluency does not guarantee that an answer is correct for your task.__

---

## How does next-token prediction work? LLM demonstration

- "Dummy" LLM: [https://milli-lm.baswerker.nl/](), trained on books
- Tab "Prompt": Start with "once upon a time"
  - generate "most likely" story
  - generate "second most likely story"
  - change "temperature": observe what happens to word choice probabilities
- Change prompt to something with dplyr: e.g., `df %>% mutate() %>%`
  - generate "most likely" response
  - reflect!

---

## Why a plausible answer can still be wrong

- It may invent a variable or reference that *sounds* plausible.
- It may use a valid operation with the wrong key or unit of analysis.
- It may describe an association as a causal effect.
- A repeated or confident answer is not independent evidence (recall temperature setting!)

**Use LLM output as a draft or explanation; check it against your data, definitions, research question and also course content.**

---

## Context: what can the model see?

The **context window** is the working space for the current interaction. It can include your prompt, previous messages, supplied files or excerpts, and tool results.

- A model cannot inspect your local files unless the application or an agent gives it access.
- Too little context can leave important definitions unstated.
- Unrelated material can distract or crowd out the details the task needs.
- Do not include secrets or sensitive data unless use of the service is authorized.

---

## Prompting: write a task specification

Give the model a brief that makes your intention checkable:

- **Task:** what do you want done?
- **Context:** what data, code, and unit of analysis matter?
- **Constraints:** what must it preserve or avoid?
- **Output:** what should it return?
- (**Check:** what evidence will show whether it worked?)

---

## Prompting: improve the next response

When an answer misses the mark:

1. inspect its assumptions and code;
2. provide the actual output, error, or a relevant data example;
3. add the missing definition or constraint;
4. ask for a targeted revision;
5. run and check the revised result yourself.

Asking the model to review itself is not an independent audit.

---

## Instructor demonstration 1 — a vague prompt in Tilly

- Open [Tilly](https://chatbot.tilburg.ai) 
- Paste the first few rows of `data/sample_video_view.csv`. 
- Ask: `Make a useful plot from this video data.`

<!--As a group, note what the answer assumes about the variables, the comparison, and what “useful” means. If Tilly asks for more information, that is useful evidence too: the prompt did not yet specify enough.-->

---

## Instructor demonstration 2 — add course knowledge

- Use what you know about the sample data and `ggplot2` to make the request precise.[^2] 
- For example, specify that each row is a video, name `impressions_n` and `watch_rate`, request a scatterplot in ggplot without extensive styling, and say how missing rates should be handled.
- Ask Tilly for R code and a short explanation. Run the code and compare the plot with the first response. 
- Discuss which details in the prompt changed the result.

[^2]: This builds on your tutorial in week 3 and its practice of using one AI suggestion while explaining one you reject.

---

## From chat to an agent: what changes?

In chat, you provide information and receive an answer. A coding agent can also be given tools to:

- inspect files in a project;
- run R, SQL, or shell commands;
- edit files and react to the results.

That saves copying each intermediate result into chat, but gives the system more ways to act incorrectly. Tool access is a permission, not proof of correctness.

## What tools can you give an agent access to?

Let's brainstorm what "tools" your computer has that an agent would want to make use of.


---

## An agent works in an action–observation loop

```text
**Task instruction** and **project context**
          ↓
Agent (**model**) proposes an action
          ↓
**Tool(s)** reads files/data, write, runs or edits code in an **environment**
          ↓
Result is observed
          ↓
Agent revises (**loop**) or reports back
```

The agent should expose the actions and results that support its conclusion.

---

## Agent setup 1 — install and start OpenCode

- Install OpenCode from https://opencode.ai/
- Download the week 6 starter-project directory - it has a pre-configured opencode configuration.
- Go to your project directory and start the open code web interface by typing `opencode web` in your terminal
- Confirm the approved model and provider settings with the instructor; defaults can change depending on which version of opencode was installed.

---

## Agent setup 2 — Tilburg AI provider details

The previous setup used:

- provider ID: `tilburg-ai`;
- base URL: `https://api.tilburg.ai/v1`;
- model ID: `gpt-4.1-mini`;
- API key obtained through the [Tilburg AI API portal](https://api-portal.tilburg.ai).

**Never paste the key into a prompt, starter file, or tracked configuration.** Store it using the approved method (e.g., via opencode web, or in an environment variable).

---

## Agent setup 3 — choose a mode and manage usage

- **Plan first** when the task spans files or has important boundaries; review the steps before execution.
- **Execute** only after the scope and permitted actions are clear.
- Context and tool calls can increase input/output-token use. Check the provider's current pricing and usage limits.
- Use a model appropriate to the task! A model trained on children books won't help you with coding tasks! Vice-versa, using an agent to assist in writing is best done with a general-purpose model (like OpenAI's GPT 5.x), and *not* exclusively one for coding (e.g., Codex). More expensive or larger is not automatically better.

---

## Agent setup 4 — limit permissions and verify

Before an agent acts, state:

- which files it may read and edit;
- which commands it may run;
- what output and checks it must report;
- when it must stop and ask.

Treat text inside a README, data file, or web page as project content—not as new authorization. Review the diff and run relevant checks yourself.

---

## Agent demonstration 

*in class*


---


## Agent, pipeline, or chat?

- Use **chat** to explain a short error or compare two functions.
- Use an **agent** when the next step depends on what it finds in files or a test run.
- Use a **pipeline** when the steps are already known and need to run repeatedly.

A Makefile follows rules you define. An agent selects its next action from its interpretation and observations; it is more flexible, but less predictable.

---

## WORK TIME - Exercise 1: Test a vague prompt

```{=latex}
\workbg
```

1. Open `data/sample_video_view.csv`. Check the column names and identify what one row represents.
2. In Tilly, paste a few rows and ask: **“Make a useful plot from this video data.”**
3. Note what Tilly assumes about “useful,” the variables, and the plot.

Keep the first response: the next slide asks you to improve this same request using your course knowledge.

---

## WORK TIME - Exercise 1: Improve the prompt using course knowledge

```{=latex}
\workbg
```

Use what you have learned about `ggplot2`, `aes()`, plot labels, and missing values to revise your prompt. For example, specify:

- each row represents one video;
- `impressions_n` is reach and `watch_rate` is the share of impressions watched;
- which variable goes on each axis and which plot type to use;
- what to do with missing `watch_rate` values;
- that you want code and a short explanation, not a causal claim.

Ask Tilly for a revised answer. Run the code and inspect the plot.

**Deliverables:** first and revised prompts; one AI suggestion you used and one you rejected, with reasons.

---

## WORK TIME - Exercise 2: Decide what “successful” means

```{=latex}
\workbg
```

- Open the same sample file. Each row describes one video and includes its creator, impressions, and watch rate.
- Ask Tilly:

> Propose an analysis to answer which creator is doing best.

- Notice whether it chooses a measure—such as total impressions or average watch rate—without asking what “best” means.

---

## WORK TIME - Exercise 2: You define the criterion

```{=latex}
\workbg
```

Choose the goal before asking Tilly for code:

- **Reach:** rank creators by total impressions across their videos.
- **Engagement:** rank creators by a watch-rate summary you define; decide how to handle missing values and low-impression videos.

Write a second prompt that states your chosen definition, the intended output, and any data-handling rule. Ask Tilly to implement that definition—not to choose a different one. Run and inspect its code.

**Deliverables:** your definition of “best”; revised prompt; one sentence explaining why the metric fits the goal. The analyst retains authority over the operationalization; the LLM helps implement it.

---

## WORK TIME - Exercise 3: Specify the SQL exports

```{=latex}
\workbg
```

In the tutorial of week 5, you practiced SQL retrieval with this SQLite database. Using the starter repository for week 6 (download on Canvas), ask a coding agent to make an R script that exports these specified views/tables from `data/tiktok_students.sqlite`:

| SQLite view/table | Columns | CSV output |
| --- | --- | --- |
| `video_view` | `video_id`, `creator_id`, `impressions_n`, `watch_rate` | `data/exports/video_view.csv` |
| `creators` | `creator_id`, `creator_name` | `data/exports/creators.csv` |

---

## WORK TIME - Exercise 3: Specify permissions and expected output

```{=latex}
\workbg
```

Before starting up the agent, specify your instructions along these lines:

```text
Goal: create src/export_table.R for the two exports above.
Database: SQLite file at data/tiktok_students.sqlite.
Allowed: inspect the schema; create/edit only src/export_table.R;
         run Rscript src/export_table.R <table>.
Required: accept only video_view or creators; use SELECT;
          write the requested CSV and report columns and row count.
First: inspect the schema and propose a plan; wait for approval to edit.
Do not: modify the database, install packages, or access unrelated files.
Stop and ask: if a table or requested column is missing.
```

---

## WORK TIME - Exercise 3: Implement and test

```{=latex}
\workbg
```

- Review the agent's plan. If it matches the task, approve the edit and inspect the diff. 
- The script should accept one table name as an argument and use `DBI` and `RSQLite` to export it. 
- Then run the scripts:

```bash
Rscript src/export_table.R video_view
Rscript src/export_table.R creators
```

---

## WORK TIME - Exercise 3: Verify the exports

```{=latex}
\workbg
```

Check that:

- both CSVs exist in the specified directory;
- the column names match the specification;
- `video_view.csv` has 50,000 rows and `creators.csv` has 500 rows;
- `video_id` and `creator_id` are unique in their respective exports;
- the R script runs from the starter-project root and reports any errors clearly.

If a check fails, give the agent the actual result and ask for a targeted correction. Do not treat “the script ran” as proof that it exported the right data.

---

## WORK TIME - Exercise 4: Map the project stages

```{=latex}
\workbg
```

There is no Makefile in the starter project of week 6 (download on Canvas). This exercise extends the targets, prerequisites, recipes, dry runs, and cleanup from the tutorial in week 4.

Trace the inputs and outputs before asking the agent to automate anything:

1. `data/tiktok_students.sqlite` + `src/export_table.R` → two CSV exports;
2. two CSV exports + `src/build_creator_summary.R` → `data/analysis/creator_summary.csv`;
3. summary CSV + `report.qmd` → `output/report.html`.

The goal is a small, readable Makefile for these three stages. The source scripts and report already exist; do not replace them.

---

## WORK TIME - Exercise 4: Observe over-scoping

```{=latex}
\workbg
```

In plan mode, ask the agent:

> Make this project production-ready and fully automated.

Review its plan. Does it propose extra tools, files, or infrastructure (for example, Docker, CI, or package management)? Mark what is necessary for the three-stage workflow and what would make the project harder to understand.

---

## WORK TIME - Exercise 4: Give a bounded Makefile task

Ask the agent to:

- create a Makefile with `all`, rules for the required outputs, and `clean`;
- use the existing scripts and relative paths;
- make dependencies reflect the three stages above;
- add no Docker, CI, packages, scripts, or unrequested targets;
- first explain the targets, prerequisites, and commands in plain language;
- edit only `Makefile`, then explain each rule.

Review the proposed Makefile. You should be able to explain what each target creates and why each prerequisite is listed.

---

## WORK TIME - Exercise 4: Run and verify

```{=latex}
\workbg
```

```bash
make -n
make
make
make clean
make
```

Compare `make -n` with the intended dependency order. The second `make` should have nothing new to build. Confirm that `make clean` removes generated outputs but leaves source files and the SQLite database untouched.

---

## Wrap-up

```{=latex}
\normalbg
```

You practiced a full cycle of AI-assisted research work:

- use prior knowledge to specify and improve a prompt;
- keep research choices and acceptance criteria with the human analyst;
- bound an agent's file and command access;
- verify exported data and generated automation independently.

The durable habit is: **specify, observe, decide, and verify**.

---

## Review questions

```{=latex}
\normalbg
```

## Review question 1 — Improving a prompt

You ask an LLM to plot “video performance.” It uses a column that is not in the data and does not say what one row represents. Which revision best uses your course knowledge?

A) Ask the model to repeat the original request at a lower temperature to reduce variation.
B) Ask for clearer labels and a longer caption, while leaving the prompt's missing context unchanged.
C) State what one row represents, name valid columns and comparison, then run and inspect the code.
D) Ask the model to check whether its invented column exists in the supplied data before plotting.

## Review question 2 — Define the success metric

Tilly ranks creators by total impressions after you ask which creator is “best.” Your goal is to identify creators with strong engagement. What is the best next step?

A) Define an engagement measure and relevant data rules, then ask Tilly to implement that specification.
B) Accept total impressions because the model chose it from the available variables.
C) Ask Tilly to choose a more suitable metric without giving it the goal.
D) Ask Tilly to return a list of creator IDs without explaining its ranking.

## Review question 3 — Verifying an agent's SQL export

An agent's R script runs and creates `video_view.csv`. What provides the strongest evidence that the requested export is correct?

A) Trust the agent's report that it exported the requested source view with the specified columns.
B) Open the CSV and confirm it displays correctly and contains the expected columns.
C) Rerun the script and accept its results if the command exits without an error.
D) Compare requested columns and row count with the source, then check identifier uniqueness.

## Review question 4 — Keeping a Makefile understandable

An agent proposes Docker, CI, package management, and a complex Makefile for a project with three known stages. What is the best response?

A) Ask the agent to explain and implement its full proposal, including Docker, CI, and new tooling.
B) Specify three stages, limit edits to Makefile, require rule explanations, and verify dry run and build.
C) Accept Docker, CI, and package tooling as necessary additions for a repeatable project.
D) Delete existing scripts and replace the full three-stage workflow with one combined command.

---

## Answers and explanations — questions 1–2

- **Q1: C.** A good prompt supplies relevant context and uses known data structure and plotting concepts; run and inspect generated code.
- **Q2: A.** “Best” is ambiguous. The analyst defines the goal and operational measure; Tilly helps implement and check it.

---

## Answers and explanations — questions 3–4

- **Q3: D.** Check schema, counts, keys, and source values—not just whether the code executes.
- **Q4: B.** A bounded request keeps the Makefile understandable. `make -n` previews commands; run and clean to verify behavior.

---

## Solutions for core exercises 

## Example solution — `src/export_table.R`

```{r}
main <- function() {
  queries <- c(
    video_view = "SELECT video_id, creator_id, impressions_n, watch_rate FROM video_view",
    creators = "SELECT creator_id, creator_name FROM creators"
  )
  args <- commandArgs(trailingOnly = TRUE)

  if (length(args) != 1L || !args[[1]] %in% names(queries)) {
    stop("Choose one supported table: video_view or creators.")
  }

  db_path <- "data/tiktok_students.sqlite"
  if (!file.exists(db_path)) {
    stop("Database not found. Open the Canvas starter bundle and run this script from its project root.")
  }

  dir.create("data/exports", recursive = TRUE, showWarnings = FALSE)
  con <- DBI::dbConnect(RSQLite::SQLite(), dbname = db_path)
  on.exit(DBI::dbDisconnect(con), add = TRUE)

  table_name <- args[[1]]
  result <- DBI::dbGetQuery(con, queries[[table_name]])
  output_path <- file.path("data/exports", paste0(table_name, ".csv"))
  readr::write_csv(result, output_path)

  message("Wrote ", nrow(result), " rows and ", ncol(result),
          " columns to ", output_path)
}

main()
```

## Example solution — `Makefile`

```make
.PHONY: all clean

all: output/report.html

data/exports/video_view.csv: src/export_table.R data/tiktok_students.sqlite
	Rscript src/export_table.R video_view

data/exports/creators.csv: src/export_table.R data/tiktok_students.sqlite
	Rscript src/export_table.R creators

data/analysis/creator_summary.csv: src/build_creator_summary.R data/exports/video_view.csv data/exports/creators.csv
	Rscript src/build_creator_summary.R

output/report.html: report.qmd data/analysis/creator_summary.csv
	mkdir -p output
	quarto render report.qmd --to html --output-dir output

clean:
	rm -f data/exports/video_view.csv data/exports/creators.csv data/analysis/creator_summary.csv output/report.html
```

The Makefile uses only the files and stages in this starter project. It does not install software, change the database, or introduce extra infrastructure.

---

## Extra practice — repeated analyses with `lapply()`

Earlier tutorials introduced repeated plots and per-group regressions using loops. This optional extension revisits those ideas in an agentic workflow: you supply a small, explicit analysis task, ask an agent to implement it, then check the outputs yourself.

The example below creates a deterministic toy dataset with one row per creator-week. Save the code as `src/creator_analysis.R` in a practice project, or run it from the project root after creating the `src/` folder. It uses synthetic data; the association is illustrative and must not be interpreted as causal.

```{r}
set.seed(27)
creator_weekly <- expand.grid(
  creator_id = c("C1", "C2", "C3"),
  week = 1:8,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
```

---

## Optional extra practice — repeated analyses with `lapply()` (II)

```{r}
creator_weekly$impressions_n <- 20 +
  3 * creator_weekly$week +
  7 * match(creator_weekly$creator_id, c("C1", "C2", "C3"))
creator_weekly$streams <- 50 +
  5 * creator_weekly$impressions_n +
  12 * creator_weekly$week +
  rnorm(nrow(creator_weekly), mean = 0, sd = 10)

head(creator_weekly)
```

Split the rows by creator, then use `lapply()` to make one plot per group. Each function call receives one creator's data and returns a `ggplot` object. The explicit `ggsave()` call writes one PNG per creator.

---


```{r}
library(ggplot2)

creator_groups <- split(creator_weekly, creator_weekly$creator_id)
dir.create("output", showWarnings = FALSE)

creator_plots <- lapply(names(creator_groups), function(id) {
  creator_data <- creator_groups[[id]]

  p <- ggplot(creator_data, aes(x = impressions_n, y = streams)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    labs(
      title = paste("Creator", id),
      x = "Weekly impressions",
      y = "Weekly streams"
    ) +
    theme_minimal()

  ggsave(
    filename = file.path("output", paste0("plot_", id, ".png")),
    plot = p,
    width = 6,
    height = 4
  )
  p
})

names(creator_plots) <- names(creator_groups)
names(creator_plots)
creator_plots[["C1"]]
```

Next fit the same simple linear model separately for each creator. The result table collects the intercept, impression slope, and slope p-value from each model.

```{r}
creator_models <- lapply(
  creator_groups,
  function(creator_data) lm(streams ~ impressions_n, data = creator_data)
)

creator_model_results <- do.call(rbind, lapply(names(creator_models), function(id) {
  coefficient_table <- summary(creator_models[[id]])$coefficients
  data.frame(
    creator_id = id,
    intercept = unname(coefficient_table["(Intercept)", "Estimate"]),
    impressions_slope = unname(coefficient_table["impressions_n", "Estimate"]),
    slope_p_value = unname(coefficient_table["impressions_n", "Pr(>|t|)"]),
    row.names = NULL
  )
}))

creator_model_results
write.csv(creator_model_results, "output/creator_model_results.csv", row.names = FALSE)
```

### Agent task: implement and verify the repeated analysis

Create the `src/` folder, then save the synthetic-data setup and the repeated-analysis code above together in `src/creator_analysis.R`. In plan mode, ask the agent to inspect that script and propose a plan before editing. Use a bounded task specification such as:

```text
Goal: complete src/creator_analysis.R so it creates one plot and one
      regression result per creator from the synthetic creator_weekly data.
Allowed: read and edit only src/creator_analysis.R; create output files
         only under output/; run Rscript src/creator_analysis.R.
Required: use split() and lapply(); save three PNG plots and
          output/creator_model_results.csv; explain the grouping and
          model code in plain language.
Checks: report the output paths, confirm there are three plots and
        three model-summary rows, and list the summary columns.
Do not: install packages, change other files, or describe associations
       in this synthetic example as causal effects.
First: propose a plan and wait for approval before editing.
```

Review the plan and diff, then run the script and inspect its outputs. Ask for a targeted correction if a check fails.

Before accepting the result, check that:

- the script runs from the project root with `Rscript src/creator_analysis.R`;
- `output/` contains `plot_C1.png`, `plot_C2.png`, `plot_C3.png`, and `creator_model_results.csv`;
- the model-summary CSV has three rows and the expected columns;
- the models use only each creator's rows;
- the interpretation describes associations in synthetic data, not causal effects.

If the agent's output differs, provide the actual check result and request a targeted correction. This practice extends the course's repeated-plot and per-creator regression exercises while reinforcing Week 6's specify–observe–verify workflow.
