---
title: "Data Challenge 1"
author: "YOUR NAME"
date: "DATE"
output: html_document
---

We're going to run our analysis in a R Markdown file which is somewhat similar to a Jupyter Notebook often used for Python scripts. Both tools allow you to combine inline code snippets (the part between ```{r}```) with text in markdown (e.g., `#### Exercise 1`). For now, it's important to realize that once you click on the "Knit" button, this `.Rmd` file is rendered into a HTML page. As you'll see, this creates headers, separates markdown cells from code blocks, makes links clickable, and many other things. Have a watch at this video [tutorial](https://www.youtube.com/watch?v=tKUufzpoHDE) or play round with this [cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf) to learn the basics. 

Please keep in mind that your notebook only runs if there are no errors. In case you suspect there might be an error somewhere, you may want to add `eval=FALSE` to `{r}` so that it skips that block and still renders your file.

\

#### Exercise 1

```{r}
mobility <- read.csv("2020_NL_Region_Mobility_Report.csv", sep=",")
# your answer
```

\ 

#### Exercise 2

```{r}
# your answer
```

\ 

#### Exercise 3

```{r}
# your answer
```

\

#### Exercise 4

```{r}
# your answer
```

\

#### Exercise 5

```{r}
# your answer
```

\

#### Exercise 6

```{r}
# your answer
```

\

#### Exercise 7

```{r}
# your answer
```

\

#### Exercise 8

```{r}
# your answer
```

\

#### Analysis & write-up of open assignment

```{r}
# your answer
```












